Astr 513: Final Project

### 2016-05-05 Final_Project/ms.tex
Modified intro and distance mod sections a bit. Added paragraph to results section.

\section{Introduction}
\label{sec:Intro}
Type \rom{1}a supernovae (\snea{}) have been instrumental in the discovery of the accelerating expansion of the universe. Presently, {\sn} surveys continue to be pivotal in constraining cosmological parameters. In this paper, we present our analysis of comparing and combining {\sna} inference to those from the Cosmic Microwave Background (CMB) and the Baryonic Acoustic Oscillations (BAO). The constraints from these two probes are taken from \citet{planck2013}. This paper is outlined as follows. In Section 2, the process and importance of {\snea} are summarized. A short description of the Planck mission is given in Section 3. Section 4 describe the two {\sna} surveys whose data we use for our analysis. The significance of using distance modulus data to constrain cosmological parameters is illustrated in Section 5. The CMB and BAO priors taken from \citet{planck2013} are also specified in Section 5. Our results and analysis on the benefit of omitting or including Bayesian priors from other cosmological probes are in Section 6 and Section 7.

\section{Type \rom{1}a Supernovae}
\label{sec:SNIa}
A type \rom{1}a supernova (\sna{}) is a supernova with very little hydrogen or
helium absorption. It is believed that the progenitor for these special
supernovae are white dwarfs that are part of a binary system. These white
dwarfs accrete mass from their companion stars or merge with a companion white
dwarf or neutron star, causing them to exceed the Chandrasekhar limit
($1.44 M_\odot$). When this happens, carbon and oxygen fusion begins in the
core, and the temperature increases. However, the white dwarf does not expand
and cool as a main sequence star would, since it is supported by degeneracy
pressure rather than thermal, so the core experiences a runaway fusion
reaction, releasing $\sim 10^{44}$ J of energy \citep[see, e.g.,][chap. 18.4]{ryden2010}, 
which is enough to overcome the gravitational energy and the star explodes. Because all 
\snea{} are caused by a similar reaction, they all have similar luminosities, and more 
importantly, the light curves all have the same shape: a sharp maximum followed by a smooth, 
gradual decay. Studying nearby \snea{} has shown that the rate of decline of the light curve 
is correlated to the peak luminosity, and this correlation has been calibrated using Cephied 
variable stars located in the same galaxies as nearby \sna{}. With this information, one can 
estimate peak luminosity of an \sna{} from its light curve, thereby providing a 
standard candle with which to measure distance. They were first used in this way in the 
1990s by two independent teams, the Supernova Cosmology Project and the Cal\'{a}n/Tololo 
Supernova Survey. It was using distances from \snea{} from these two projects that 
\citet{riess1998} and \citet{perlmutter1999} were able to discover the accelerating expansion 
rate of the Universe in 1998.

\section{Planck 2015 Results}
\label{sec:planck}
In 2009 the European Space Agency launched the \Planck satellite in order
to measure the Cosmis Microwave Background (CMB) with unparalleled precision.
By measuring the small anisotropies in the temperature of the CMB across
the entire sky one calculates the angular power spectrum, which can be
modeled using first principles (see, e.g., \citet{dodelson2003}).
By using a likelihood method similar to our own, the authors of
\citet{planck2013} estimated 14 nuisance parameters and 
six cosmological parameters, including the geometrical parameters,
$\Omega_M$, $\Omega_\Lambda$ and $\Omega_k$, that we use in our
{\sna} analysis. The only prior used by \citet{planck2013}
on cosmological parameters was from the WMAP result \citep{wmap2011}
for the optical depth $\tau$. We do not use this parameter
in our {\sna} model.

@@

\section{Distance Modulus Model}
\label{sec:model}
The main objective of \citet{betoule2014} and \citet{rest2014} is 
to place constraints on cosmological parameters $\Omega_M$, $\Omega_\Lambda$, 
$\Omega_k$, and $w$. In order to so, we must examine the relationship between distance 
modulus and cosmology. The distance modulus $\mu$ is determined from 
the luminosity distance $d_L$ by 
\begin{equation}
\label{eq:mu_dL}
\mu=5 \log\left(\frac{d_L}{1 \mathrm{Mpc}}\right)+25. 
\end{equation}
The luminosity distance depends on the cosmology by 
\begin{equation}
\label{eq:luminosityDistance}
d_L(z; w, \Omega_M, \Omega_\Lambda, H_0)= (1+z)|\Omega_k|^{-1/2}\mathcal{S}_{k}\displaystyle\left[c|\Omega_k|^{1/2}\int_{0}^{z}\frac{dz'}{H(z')}\right],
\end{equation}
where $\Omega_k \equiv 1-\Omega_M-\Omega_{\Lambda}$. The function 
$\mathcal{S}_{k}$ depends on the curvature density $\Omega_k$ in that 
\[
    \mathcal{S}_{k}= 
\begin{cases}
    \text{sin}(x),&  \Omega_k < 0\\
    x,& \Omega_k = 0\\
    \text{sinh}(x),& \Omega_k > 0
\end{cases}
\]
and the function $H(z)$ is 
\begin{equation}
\label{eq:hubbleFunction}
H(z)= H_0\displaystyle\left[\Omega_M(1+z)^3+\Omega_{\Lambda}(1+z)^{3(1+w)}+\Omega_k(1+z)^2\right]^{1/2}.
\end{equation}
Two cosmologies are of interest to us: a flat universe where $\Omega_k=0$ and a universe where $\Omega_k$ is unconstrained.

@@

\section{Results}
\label{sec:results}
\TM{Discuss the results and show the corner plots. Example plot included.
  We should have one corner plot without priors and one with priors.}

>>>>

\section{Introduction}
\label{sec:Intro}
Type \rom{1}a supernovae (\snea{}) have been instrumental in the discovery of the accelerating expansion of the universe. Presently, {\sn} surveys continue to be pivotal in constraining cosmological parameters. In this paper, we present our analysis of comparing and combining {\sna} inference of cosmological parameters to those from other cosmological probes. In particular, we utilize the constraints from the Cosmic Microwave Background (CMB) and the Baryonic Acoustic Oscillations (BAO) that are gathered in \citet{planck2013}. We look at two cosmologies in our analysis: the $\Lambda \textnormal{CDM}$ universe and the wCDM universe. This paper is outlined as follows. In Section 2, the process and importance of {\snea} are summarized. A short description of the Planck mission is given in Section 3. Section 4 describe the two {\sna} surveys whose data we use for our analysis. The significance of using distance modulus data to constrain cosmological parameters is illustrated in Section 5. The CMB and BAO priors taken from \citet{planck2013} are also specified in Section 5. Our results and analysis on the benefit of omitting or including Bayesian priors from other cosmological probes are in Section 6 and Section 7.

\section{Type \rom{1}a Supernovae}
\label{sec:SNIa}
A type \rom{1}a supernova is a supernova with very little hydrogen or
helium absorption. It is believed that the progenitor for these special
supernovae are white dwarfs that are part of a binary system. These white
dwarfs accrete mass from their companion stars or merge with a companion white
dwarf or neutron star, causing them to exceed the Chandrasekhar limit
($1.44 M_\odot$). When this happens, carbon and oxygen fusion begins in the
core, and the temperature increases. However, the white dwarf does not expand
and cool as a main sequence star would, since it is supported by degeneracy
pressure rather than thermal, so the core experiences a runaway fusion
reaction, releasing $\sim 10^{44}$ J of energy \citep[see, e.g.,][chap. 18.4]{ryden2010}, 
which is enough to overcome the gravitational energy and the star explodes. Because all 
\snea{} are caused by a similar reaction, they all have similar luminosities, and more 
importantly, the light curves all have the same shape: a sharp maximum followed by a smooth, 
gradual decay. Studying nearby \snea{} has shown that the rate of decline of the light curve 
is correlated to the peak luminosity, and this correlation has been calibrated using Cephied 
variable stars located in the same galaxies as nearby \sna{}. With this information, one can 
estimate peak luminosity of an \sna{} from its light curve, thereby providing a 
standard candle with which to measure distance. They were first used in this way in the 
1990s by two independent teams, the Supernova Cosmology Project and the Cal\'{a}n/Tololo 
Supernova Survey. It was using distances from \snea{} from these two projects that 
\citet{riess1998} and \citet{perlmutter1999} were able to discover the accelerating expansion 
rate of the Universe in 1998.

\section{Planck 2015 Results}
\label{sec:planck}
In 2009 the European Space Agency launched the \Planck satellite in order
to measure the Cosmis Microwave Background with unparalleled precision.
By measuring the small anisotropies in the temperature of the CMB across
the entire sky one calculates the angular power spectrum, which can be
modeled using first principles (see, e.g., \citet{dodelson2003}).
By using a likelihood method similar to our own, the authors of
\citet{planck2013} estimated 14 nuisance parameters and 
six cosmological parameters, including the geometrical parameters,
$\Omega_M$, $\Omega_\Lambda$ and $\Omega_k$, that we use in our
{\sna} analysis. The only prior used by \citet{planck2013}
on cosmological parameters was from the WMAP result \citep{wmap2011}
for the optical depth $\tau$. We do not use this parameter
in our {\sna} model.

@@

\section{Distance Modulus Model}
\label{sec:model}
The main objective of \citet{betoule2014} and \citet{rest2014} is 
to place constraints on cosmological parameters $\Omega_M$, $\Omega_\Lambda$, 
$\Omega_k$, and $w$. In order to so, we must examine the relationship between distance 
modulus and cosmology. The distance modulus $\mu$ is determined from 
the luminosity distance $d_L$ by 
\begin{equation}
\label{eq:mu_dL}
\mu=5 \log\left(\frac{d_L}{1 \mathrm{Mpc}}\right)+25. 
\end{equation}
The luminosity distance depends on the cosmology by 
\begin{equation}
\label{eq:luminosityDistance}
d_L(z; w, \Omega_M, \Omega_\Lambda, H_0)= (1+z)|\Omega_k|^{-1/2}\mathcal{S}_{k}\displaystyle\left[c|\Omega_k|^{1/2}\int_{0}^{z}\frac{dz'}{H(z')}\right],
\end{equation}
where $\Omega_k \equiv 1-\Omega_M-\Omega_{\Lambda}$. The function 
$\mathcal{S}_{k}$ depends on the curvature density $\Omega_k$ in that 
\[
    \mathcal{S}_{k}= 
\begin{cases}
    \text{sin}(x),&  \Omega_k < 0\\
    x,& \Omega_k = 0\\
    \text{sinh}(x),& \Omega_k > 0
\end{cases}.
\]
The function $H(z)$ is expressed as 
\begin{equation}
\label{eq:hubbleFunction}
H(z)= H_0\displaystyle\left[\Omega_M(1+z)^3+\Omega_{\Lambda}(1+z)^{3(1+w)}+\Omega_k(1+z)^2\right]^{1/2},
\end{equation} where the Hubble constant $H_0$ can be written in terms of a dimensionless Hubble parameter $h$ by $H_0=100h \mathrm{kms^{-1}Mpc^{-1}}$.

Two cosmologies are of interest to us: the $\Lambda \textnormal{CDM}$ universe and the wCDM universe. In the $\Lambda \textnormal{CDM}$ universe, $w=-1$ and constraints are placed on $\Omega_M$ and $h$. The wCDM universe is a one-parameter extension of $\Lambda \textnormal{CDM}$ in which $w$ is an arbitrary parameter for a constant dark energy equation of state. In wCDM, constraints are placed on $\Omega_M$, $w$, and $h$. In both cosmologies, the universe is spatially flat ($\Omega_k = 0$) and thus $\Omega_{\Lambda}=1-\Omega_M$.

@@

\section{Results}
\label{sec:results}
\TM{Discuss the results and show the corner plots. Example plot included.
  We should have one corner plot without priors and one with priors.}
\LC{Not sure where to put why we couldn't do the full analysis of Rest's data. I'm putting it here for now.}
A complete analysis using the data from \citet{rest2014} is not done in our paper since there are several issues with the publicly released data. It is not apparent which uncertainties belonged to the statistical covariance matrix since only the systematic covariance matrix is clearly presented. Moreover, intrinsic dispersion of the data and photometric errors of the distances depend on nuisance parameters in addition to the covariances between the fit parameters. Fine-tuning nuisance parameters places a significant time restraint on our work and is beyond the scope of this paper. For these reasons, we only present the results for \citet{rest2014} that do not include the priors from \citet{planck2013}.
